---
title: "assignment_2_plotting_s.Rmd"
author: "Serban Aliz"
date: "2024-11-23"
output: html_document
chunk_output_type: console
---
You will have to create 3 plots based on the datasets and instructions detailed below. You will find the plots themeselves in the `assignments/assignment_2_plots`. Your task is to write the code that will reproduce the plots as closely as possible.

# Skills needed to solve this assignment

-   Using R and RStudio, reading data
-   Reporting using RMarkdown
-   Using Git and Github (for submitting the task)
-   Data manipulation (e.g. dplyr, tidyr), and working with factors (forcats)
-   Data visuzlization (ggplot2)


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(tidytuesdayR)
```


## Task 1: Climbing expeditions

The 2020-09-22 TidyTueday datasets are about climbing expeditions. From the three datasets, use the "expeditions". Reproduce the plot below! Notice a few things:

-   Use `forcats::fct_lump()` to get the 15 most frequent peaks, and drop the "Other" category.
-   The bars are ordered by the sum of all expeditions (use `fct_reorder()`).
-   The bar colors use the viridis palette and light theme.

```{r}
# load the data
tt_output <- tt_load("2020-09-22")
expeditions <- tt_output$expeditions

expeditions %>%
  # get the 15 most frequent peaks
  mutate(peaks = forcats::fct_lump(expeditions$peak_name, n = 15)) %>%
  # drop the "Other" category
  filter(peaks != "Other") %>%
  # for each peak and season
  group_by(peaks, season) %>%
  # find the number of expeditions
  summarise(n_expeditions = n()) %>%
  # find total expeditions for each peak
  mutate(total_expeditions = sum(n_expeditions)) %>%
  # create a barplot
  ggplot(aes(x = fct_reorder(peaks, total_expeditions),
             y = n_expeditions,
             fill = season)) +
  geom_col() +
  # make the plot horizontal
  coord_flip() +
  # change bar colors
  scale_fill_viridis_d() +
  # change to ligt theme
  theme_light() +
  # move legend to bottom
  theme(legend.position = "bottom") +
  # change axis labels and add a title
  labs(y = "Number of expeditions",
       x = "",
       title = "The 15 most popular peaks stacked by season of expedition")
```

## Task 2: PhDs awarded

The 2019-02-19 TidyTueday dataset is about phd-s awarded by year and field. There is only one dataset, it is called `phd_by_field`. Reproduce the plot below!

Notes:

-   First you have to aggregate the data to count all phd-s by broad fields.
-   To make the x axis breaks pretty, use `scales::pretty_breaks()`, to make the y axis labels comma formatted, use `scales::comma_format()`.
-   The line size is 1.2, the colors are from the brewer "Dark2" palette. The theme is set to minimal.

```{r}
# load the data
tt_output <- tt_load("2019-02-19")
phd_by_field <- tt_output$phd_by_field

phd_by_field %>%
  # aggregate the data to count all phd-s by broad fields and year
  group_by(broad_field, year) %>%
  summarise(n_phds = sum(n_phds, na.rm = TRUE)) %>%
  # create a line plot
  ggplot(aes(x = year, y = n_phds, col = broad_field)) +
  geom_line(linewidth = 1.2) +
  # make the x axis breaks pretty
  scale_x_continuous(breaks = scales::pretty_breaks()) +
  # make the y axis labels comma formatted
  scale_y_continuous(labels = scales::comma_format()) +
  # change color palette
  scale_color_brewer(palette = "Dark2") +
  # change theme to minimal
  theme_minimal() +
  # change axis labels and add a title
  labs(x = "",
       y = "",
       col = "Broad field",
       title = "Number of awarded Ph.D.-s in the US by year")
```

## Task 3: Commute in the US

The 2019-11-05 TidyTueday dataset is about commuting to work in each city in the US by bike or on foot. There is only one dataset, it is called `commute`. Reproduce the plot below!

Notes:

-   First you have to aggregate the data to count all commutes by state.
-   Both axis scales are log transformed and the labels comma formatted, using `scales::comma_format()`
-   The point size is 2, . The theme is set to light.

```{r}
# load the data
tt_output <- tt_load("2019-11-05")
commute <- tt_output$commute

commute %>%
  # convert data to wide format
  pivot_wider(names_from = mode, values_from = n) %>%
  # aggregate the data to count all commutes by state
  group_by(state_region, state_abb) %>%
  summarise(n_walk = sum(Walk, na.rm = TRUE),
            n_bike = sum(Bike, na.rm = TRUE)) %>%
  # create a scatterplot
  ggplot(aes(x = n_walk, y = n_bike, col = state_region, label = state_abb)) +
  geom_point(size = 2) +
  geom_text(col = "black", check_overlap = TRUE) +
  # log transform axis scales and comma format the labels
  scale_x_log10(labels = scales::comma_format()) +
  scale_y_log10(labels = scales::comma_format(), limits = c(20, NA)) +
  # change theme
  theme_light() +
  # change axis labels and add a title
  labs(x = "Number of ppl walking to work (log N)",
       y = "Number of ppl biking to work (log N)",
       col = "State region",
       title = "Title number of people walking vs. biking to work in each USA state")
```
